---
title: "NEON Bird Occupancy Modeling"
author: "Anushka Gupta"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
## NEON Birds
Something something here about NEON Bird Data and how it is collected, links to protocols and data product
## Occupancy Modeling
Something something here about what occupancy modelling is and why is it useful for these data
```{r}

```

# Setup
## Importing libraries
```{r}
library(neonUtilities) #package for downloading NEON data 
library(dplyr)
library(tidyr)
```

## Importing Data
```{r}
sites_large <- ("DELA", "")
sites_small <- ()
sites_combined <- ()
bird.counts <- loadByProduct(dpID="DP1.10003.001",
                             site=c("DELA", "LENO", "TALL"),
                             startdate="2018-01",
                             enddate="2023-12",
                             check.size = F)

```

# Descriptive Analyses
## Detection
### Detection by species
Something descriptive about setup and how we're rearranging the data
```{r Setup}
the end of this should be a table with species (row) x year (cols)
```

Something descriptive about what it is we're doing
```{r Model}

```

something describing the outputs
```{r Outputs}

```

What do we see in the output? is this cool? What is it teling us? yay birds.

### Detection by species x site
### Detection by species x site x year

## Within a site
### Detection within a site
### Occupancy within a site

## Local community dynamics
### Community dynamics by species
### Community dynamics by species x site
### Community dynamics by species x site x year

# Predictive/Causual Hypotheses
