---
title: "NEON Bird Occupancy Modeling"
author: "Anushka Gupta"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
## NEON Birds
The U.S National Science Foundation's National Ecological Observatory Network is a continental-scale observation facility designed to collect long-term ecological data. 
The data product being used contains the quality-controlled, native sampling resolution data from NEON's breeding landbird sampling. 

Protocols: https://data.neonscience.org/documents/10179/1883155/NEON.DOC.014041vL/2a13bc2b-84db-e4e5-1d33-af8fac1f9e24
Data Product: https://data.neonscience.org/data-products/DP1.10003.001


## Occupancy Modeling
Occupancy models are used to help estimate true occupancy of a species and can help account for imperfect detection of organisms in a study. 

# Setup
## Importing libraries
```{r}
library(neonUtilities) #package for downloading NEON data 
library(dplyr)
library(tidyr)
```

## Importing Data
```{r}
sites_large <- c("BART","HARV","DSNY","JERC","OSBS","GUAN","STEI","UNDE","KONZ","UKFS","GRSM","MLBS","ORNL","TALL","WOOD","CPER","RMNP","CLBJ","OAES","YELL","MOAB","NIWO","JORN","SRER","ONAQ","WREF","SJER","TEAK","BARR","BONA","DEJU","HEAL")

sites_small <- c("BLAN","SCBI","SERC","LAJA","TREE","KONA","DELA","LENO","DCFS","NOGP","STER","ABBY","SOAP","TOOL","PUUM")

sites_combined <- c(large_sites, small_sites)

bird.counts <- loadByProduct(dpID="DP1.10003.001",
                             site=sites_small,
                             startdate="2018-01",
                             enddate="2023-12",
                             check.size = F)

```

# Descriptive Analyses
## Detection
### Detection by species
Something descriptive about setup and how we're rearranging the data
```{r Setup}
#the end of this should be a table with species (row) x year (cols)
counts <- bird.counts$brd_countdata
counts$year <- as.numeric(substr(counts$startDate, 1, 4)) 

# add year column
counts <- counts %>% 
          group_by(year)
View(counts)

# simplify to detection by year and species
counts_by_year <- counts %>%
  count(year, scientificName) %>% 
  filter(!is.na(scientificName))
View(counts_by_year)

# convert to wide format: species as rows, years as columns
wide_df <- counts_by_year %>% 
pivot_wider(
    id_cols = scientificName,       
    names_from = year, 
    values_from = n,
    values_fill = 0
  )

# View Resulting Table
View(wide_df)
```

Something descriptive about what it is we're doing
```{r Model}

```

something describing the outputs
```{r Outputs}

```

What do we see in the output? is this cool? What is it teling us? yay birds.

### Detection by species x site
### Detection by species x site x year

## Within a site
### Detection within a site
### Occupancy within a site

## Local community dynamics
### Community dynamics by species
### Community dynamics by species x site
### Community dynamics by species x site x year

# Predictive/Causual Hypotheses
