---
title: "NEON Bird Occupancy Modeling"
author: "Anushka Gupta"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
## NEON Birds
The U.S National Science Foundation's National Ecological Observatory Network is a continental-scale observation facility designed to collect long-term ecological data. 
The data product being used contains the quality-controlled, native sampling resolution data from NEON's breeding landbird sampling. 

Protocols: https://data.neonscience.org/documents/10179/1883155/NEON.DOC.014041vL/2a13bc2b-84db-e4e5-1d33-af8fac1f9e24
Data Product: https://data.neonscience.org/data-products/DP1.10003.001


## Occupancy Modeling
Occupancy models are used to help estimate true occupancy of a species and can help account for imperfect detection of organisms in a study. 

# Setup
## Importing libraries
```{r}
library(neonUtilities) #package for downloading NEON data 
library(dplyr)
library(tidyr)
library(stringr)
```

## Importing Data
```{r}
large_sites <- c(
  "BARR", "BART", "BONA", "CLBJ", "CPER", "DEJU", "DSNY", "GRSM",
  "GUAN", "HARV", "HEAL", "JERC", "JORN", "KONZ", "MLBS", "MOAB",
  "NIWO", "OAES", "ONAQ", "ORNL", "OSBS", "RMNP", "SJER", "SRER",
  "STEI", "TALL", "TEAK", "UKFS", "UNDE", "WOOD", "WREF", "YELL"
)
small_sites <- c(
  "DELA", "LENO", "TOOL", "SOAP", "STER", "PUUM", "KONA",
  "SERC", "DCFS", "NOGP", "LAJA", "BLAN", "SCBI", "ABBY", "TREE"
)
sites <- c(large_sites, small_sites)

if (!dir.exists("data")) {
  dir.create("data")
}

if (!dir.exists("outputs")) {
  dir.create("outputs")
}

if (file.exists("data/bird_counts.RData")) {
  load("data/bird_counts.RData")
} else {
  
  bird.counts <- loadByProduct(dpID="DP1.10003.001",
                               site=sites,
                               startdate="2018-01",
                               enddate="2023-12",
                               check.size = F)
  
  save(bird.counts, file = "data/bird_counts.RData")
}
```

## Data Examination

Describe how the data are set up (i.e., there are two tables, with what data in them?)

```{r}

```


## Data Cleanup

First, we need to do some generic data cleanup.

1. We need to filter out surveys that could not be completed for whatever reason (Covid, site was flooded, etc.)

```{r}
brd_perpoint_cleaned <- bird.counts$brd_perpoint %>%
  filter(samplingImpractical == "OK")
```

2. Since we're really only interested in species level data, we need to remove records that are not recorded to the species level

```{r}
brd_countdata_cleaned <- bird.counts$brd_countdata %>%
  filter(taxonRank == "species")
```

## Data Standardization

Because there is inconsistent sampling across sites (i.e., some sites are larger than others and not all plots within a site were continuously sampled over time), in order to do analyses across sites, we need to make sure that the data are standardized with respect to amount of sampling time.

```{r}

```

# Descriptive Analyses
## Detection
### Detection by species
Something descriptive about setup and how we're rearranging the data
```{r Setup}
#the end of this should be a table with species (row) x year (cols)
counts <- bird.counts$brd_countdata
counts$year <- as.numeric(substr(counts$startDate, 1, 4)) 

# add year column
counts <- counts %>% 
          group_by(year)
View(counts)

# simplify to detection by year and species
counts_by_year <- counts %>%
  count(year, scientificName) %>% 
  filter(!is.na(scientificName))
View(counts_by_year)

# convert to wide format: species as rows, years as columns
wide_df <- counts_by_year %>% 
pivot_wider(
    id_cols = scientificName,       
    names_from = year, 
    values_from = n,
    values_fill = 0
  )

# View Resulting Table
View(wide_df)
```

Something descriptive about what it is we're doing
```{r Model}

```

something describing the outputs
```{r Outputs}

```

What do we see in the output? is this cool? What is it teling us? yay birds.

### Detection by species x site
### Detection by species x site x year

## Within a site
### Detection within a site
### Occupancy within a site

## Local community dynamics
### Community dynamics by species
### Community dynamics by species x site
### Community dynamics by species x site x year

# Predictive/Causual Hypotheses
